---
title: "Chapter5"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Chapter 5 Data Transformation

##5.1.1  Prerequisites
```{r}
library(tidyverse)
library(nycflights13)
flights <- nycflights13::flights
head(flights)
```
##5.2 Filter rows
```{r}
filter(flights, month == 1, day == 1)
#to save selection in a new separate dataframe
jan1 <- filter(flights, month == 1, day == 1)
#to save as a dataframe and print the results
(dec25 <- filter(flights, month == 12, day == 25))
```

###5.2.1 Comparisons
```{r}
sqrt(2) ^ 2 == 2
1 / 49 * 49 == 1

near(sqrt(2) ^ 2,  2)
near(1 / 49 * 49, 1)
```

###5.2.2 Logical Operators
```{r}
#flights that departed in nov or dec
filter(flights, month == 11 | month == 12)
#to select months without running in to numerical trouble
nov_dec <- filter(flights, month %in% c(11, 12))

filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```



###5.2.4 Exercises
####5.2.4.1 flight filter exercises
```{r}
#all flights that
head(flights)
arr_two <- filter(flights, arr_delay >= 120)
hou <- filter(flights, dest == "IAH" | dest == "HOU")
op_by_del_un_am <- filter(flights, carrier == "DL" | carrier == "AA" | carrier == "UA")
sum_dep <- filter(flights, month == 7 | month == 8 | month == 9)
only_arr_late <-  filter(flights, dep_delay == 0 & arr_delay >= 120)
del_30min <- filter(flights, dep_delay >= 60 & arr_delay >= 30)

#or for del_30min
del_30min <- flights %>%
  filter(dep_delay >= 60) %>%
  mutate(difference = dep_delay-arr_delay) %>%
  filter(difference >= 30)
del_30min

flights_6am <- filter(flights, dep_time <= 600 & dep_time >= 0000)

```

####5.2.4.2 use of between
```{r}

head(flights)
arr_two_b <- between(flights$arr_delay, 2, 344)
table(is.na(flights$dep_time))



#leos example
filter(flights,between(month,7,9))

```

###5.3 Arrange
```{r}
arrange(flights, year, month, day)
arrange(flights, desc(dep_delay))

df <-  tibble(x=c(5,2,NA))
arrange(df, x)
arrange(df, desc(x))
```

####5.3.1 Exercises
```{r}
#missing values
arrange(flights, desc(is.na(arr_time)))
#most delayed flights
arrange(flights,desc(dep_delay),desc(arr_delay))
arrange(flights,dep_time)

#sort flights to find the fastest flights
arrange(flights,air_time)
arrange(flights,desc(distance))
arrange(flights,desc(distance),air_time)

#flight travelled farthest and flight travelled shortest
arrange(flights,desc(distance))
arrange(flights,distance)

```
#5.4

```{r}
#Select columns by name
select(flights, year, month, day)

#Select all columns between year and day (inclusive)
select(flights, year:day)

#Select all columns except those from year to day (inclusive)
select(flights, -(year:day))

#to move certain colums to the start of the data frame
select(flights, time_hour, air_time, everything())
```



###5.4.1 Exercises
####1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights
```{r}
head(flights)
select(flights,dep_time, dep_delay, arr_time, arr_delay)
#feels like more work or just as much
select(flights,dep_time:arr_delay,-sched_arr_time,-sched_dep_time)
```
###2. What happens if you include the name of a variable multiple times in a select() call?
```{r}
select(flights, arr_time, arr_time, arr_time)
```
###3. What does the any_of() function do? Why might it be helpful in conjunction with this vector?
```{r}
?any_of
#any of helps with selecting variables in a character vector and allows for missing values
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights,any_of(vars))
```
####4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
```{r}
select(flights, contains("TIME"))
select(flights, contains("TIME", ignore.case = TRUE))
select(flights, contains("TIME", ignore.case = FALSE))
#default is at ignore.case=TRUE

```
###5.5.2 Exercises
####1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r}
deps <- select(flights, dep_time, sched_dep_time)
deps_1 <- mutate(deps,
                 h_dep_time = dep_time %/% 100,
                 min_dep_time = dep_time %% 100,
                 totalmin_dep_time = h_dep_time*60+min_dep_time,
                 h_sch_dep_time = dep_time %/% 100,
                 min_sch_dep_time = dep_time %% 100,
                 totalmin_dep_time = h_sch_dep_time*60+min_dep_time)
deps_1

```
#### 2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?
```{r}
arrs <- mutate(flights,h_dep_time = dep_time %/% 100,
              min_dep_time = dep_time %% 100,
              totalmin_dep_time = h_dep_time*60+min_dep_time,
              h_arr_time = arr_time %/% 100,
              min_arr_time = arr_time %% 100,
              totalmin_arr_time = h_arr_time*60+min_arr_time,
              air_time2 = totalmin_arr_time - totalmin_dep_time)
arrs <- select(arrs,arr_time,dep_time,totalmin_arr_time,totalmin_dep_time,air_time,air_time2)
head(arrs)


```
#### 3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

```{r}
dep <- select(flights,dep_time,sched_dep_time,dep_delay)
dep1 <- mutate(dep,dep_delay2 = dep_time - sched_dep_time)
dep1
dep2 <- mutate(flights, h_dep_time = dep_time %/% 100,
              min_dep_time = dep_time %% 100,
              totalmin_dep_time = h_dep_time*60+min_dep_time,
              h_sched_dep_time = sched_dep_time %/% 100,
              min_sched_dep_time = sched_dep_time %% 100,
              totalmin_sched_dep_time = h_sched_dep_time*60+min_sched_dep_time,
              dep_delay2 = totalmin_dep_time - totalmin_sched_dep_time)
dep3 <- select(dep2,dep_delay,dep_delay2)
dep3
#dep3 results are in minutes
```

####4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().

```{r}
rank_delay <- mutate(flights, rank = min_rank(dep_delay))
rank_delay_desc <- arrange(rank_delay,desc(rank))
rank_delay_desc
```


####5.  What does 1:3 + 1:10 return? Why?
```{r}
1:3 + 1:10
1:3
1:10
1:10+1:3

```

####6. What trigonometric functions does R provide?
```{r}
?base::Trig
#shows all the trig functions
```

###5.6 Grouped summaries with summarise()
```{r}



```



